Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-15 paper/paper.qmd
Update to professional email
Incorporate Shaun Wilson's suggestions

authors:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
      - ref: aims
    orcid: 0009-0005-1444-6198
    attributes:
      corresponding: true

@@

abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs benefitting from reduced thermal exposure. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth should be considered as a key variable in
    conservation planning to maximize coral survival under continuing climate change.

@@

Climate change is impacting species survival, persistence and ecosystem functions on coral
reefs, which are vulnerable to increasing magnitude and variation in environmental stressors
such as ocean temperature [@foster2021; @gattuso2014]. Mass bleaching and associated
mortality events driven by climate change can cause coral population declines [@hughes2017;
@hughes2018; @bozec2025], possibly leading to coral-macroalgae regime shifts [@arif2022;
@wild2011]. Climate impacts on coral reefs are decreasing the ability of coral reefs to
provide ecological functions such as benthic habitat and nutrient cycling [@forster2021;
@williams2019] and services such as food, protection, tourism and Indigenous cultural value
[@eddy2021; @stoeckl2021].

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs (@fig-gcm-cover-trajectories). Of the GCMs explored, NorESM2-MM had the lowest ECS
(2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2075 under this
hypothetical climate future. Unsurprisingly, recovery phases were predominantly seen in the
reefs in the High cover cluster, with Medium and Low exhibiting lesser probability of such
periods of recovery. Generally, recovery and increases in coral cover were visible in
simulated results up to 2045. Under NorESM2-MM, the Far Northern Management area was found
to have the most similar temporal behaviour between Low and High clusters compared to other
Management areas (@stbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

@@

Partial Dependence Analysis revealed reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores, accounting for 32%, 16% and 15%
predictive importance respectively (@fig-ts-cluster-pdp; Panel A). Outgoing connectivity strength, Source-cover weighted
incoming connectivity and
larval retention probability account for \~14%, 14%, and 10% respectively.

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and more stable, assignment
probability across depth ranges (@fig-ts-cluster-pdp; Panel B). A large increase in the
probability of a reef being categorised as High can be seen between 10 and 15 meters depth.
This is at the edge of the depth range used in current deployment efforts [@howlett2021;
@randall2021; @whitman2024] but indicates a depth range that could be adopted to maximise
longevity and benefits of, for example, coral seeding activities. Across mean DHW values,
assignment probability is roughly equal across the three clusters up to 5 DHW. A sharp
delineation can be seen \> 5 DHW when comparing High with other clusters, with lower
assignment probability (\~25%) compared to Medium and Low (\~40%) (@fig-ts-cluster-pdp;
Panel B).

@@

First-order PDA showed minimal influence of incoming and outgoing
connectivity metrics on cluster assignment probabilities. Two-way analyses revealed little
interaction between depth and incoming and outgoing connectivity metrics for reef clustering
probabilities (@fig-ts-cluster-pdp-twoway-incoming; @sfig-ts-cluster-pdp-twoway-outgoing).
Reefs with a median depth around 15m and high incoming connectivity may benefit slightly,
with higher probability of High cover cluster assignment compared to 15m - depth reefs with
lower connectivity.

@@

Spatially these results indicate that regions of positive carbonate budget maintenance,
under ACCESS-ESM1-5, and high depth levels were concentrated in the Far Northern, offshore
Townsville/Whitsunday, and Mackay/Capricorn management areas (@sfig-depth-map;
@sfig-carb-budget-map). Resilient reefs in the Far Northern management area were exposed to
lower mean DHW levels than other reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however
similar spatial patterns were observed across GCMs for reefs clustered into the high cover
cluster (@sfig-gcm-cluster-map).

@@

The authors acknowledge the Traditional Owners of the Sea Country on which the modelling was
conducted for. The authors additionally acknowledge the Bindal, Jagera, and Ngunnawal
Peoples, the lands on which the authors conducted the model development, analysis, and
writing. The authors also acknowledge the contributions of Barbara Robson and Jessica
Benthuysen for comments provided early in the manuscript preparation. The authors would like
to additionally acknowledge and thank Shaun Wilson and Juan Carlos Ortiz for discussions
around carbonate budget thresholds.

>>>>

authors:
  - name: Benjamin Grier
    email: bengrier1@outlook.com
    affiliations:
      - ref: aims
    orcid: 0009-0005-1444-6198
    attributes:
      corresponding: true

@@

abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great
    Barrier Reef (GBR) experiencing cumulative stressors and increasingly severe declines
    in coral cover from thermal stress events. Understanding drivers behind reef resilience
    to climate impacts is critical for conservation planning and intervention strategies. A
    scenario-based population modelling approach was adopted with larval connectivity dynamics and
    environmental factors to assess coral reef resilience across the GBR using projected conditions under
    five Global Climate Models (GCMs). Projected coral
    cover was analysed for each reef's ability to maintain positive carbonate production budgets under
    future conditions, using coral cover as a proxy. Larval connectivity patterns did not correlate with
    increases in maintenance of positive carbonate budgets. Instead, reef depth emerged as the primary
    predictor, with deeper reefs (>10m) benefitting from reduced thermal exposure. These
    findings suggest that depth is a tangible and pragmatic reef characteristic to consider in future
    intervention practices for coral reef restoration. These results have important implications for reef
    management, indicating that depth should be considered as a key variable in
    conservation planning to maximize coral survival under continuing climate change.

@@

Climate change is impacting species survival, persistence and ecosystem functions on coral
reefs, which are vulnerable to increasing magnitude and variation in environmental stressors
such as ocean temperature [@foster2021; @gattuso2014]. Mass bleaching and associated
mortality events driven by climate change can cause coral population declines [@hughes2017;
@hughes2018; @bozec2025], possibly leading to coral-macroalgae regime shifts [@arif2022;
@graham2015]. Climate impacts on coral reefs are decreasing the ability of coral reefs to
provide ecological functions such as benthic habitat and nutrient cycling [@forster2021;
@williams2019] and services such as food, protection, tourism and Indigenous cultural value
[@eddy2021; @stoeckl2021].

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs (@fig-gcm-cover-trajectories). Of the GCMs explored, NorESM2-MM had the lowest ECS
(2.5°C) and reefs exhibit continuing recovery and decline dynamics up to 2075 under this
hypothetical climate future. Unsurprisingly, recovery phases were predominantly seen in the
reefs in the High cover cluster, with Medium and Low exhibiting lesser probability of such
periods of recovery. The High cover cluster exhibits greater resilience despite encompassing
a broad range of initial coral cover conditions at 2025 (\@fig-gcm-cover-trajectories).
Generally, recovery and increases in coral cover were visible in simulated results up to
2045. Under NorESM2-MM, the Far Northern Management area was found to have the most similar
temporal behaviour between Low and High clusters compared to other Management areas
(@stbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

@@

Partial Dependence Analysis revealed reef median depth, carrying capacity and mean time
series DHW level have the highest feature importance scores for bioregion-scale cluster
assignment, accounting for 32%, 16% and 15% predictive importance respectively
(@fig-ts-cluster-pdp; Panel A). Outgoing connectivity strength, Source-cover weighted
incoming connectivity and larval retention probability account for \~14%, 14%, and 10%
respectively.

Reefs have the greatest probability of being clustered into Low if shallow (\<10m), and High
at greater depths, with medium cluster reefs exhibiting a lower, and more stable, assignment
probability across depth ranges (@fig-ts-cluster-pdp; Panel B). A large increase in the
probability of a reef being categorised as High can be seen between 10 and 15 meters depth.
Across mean DHW values, assignment probability is roughly equal across the three clusters up
to 5 DHW. A sharp delineation can be seen \> 5 DHW when comparing High with other clusters,
with lower assignment probability (\~25%) compared to Medium and Low (\~40%)
(@fig-ts-cluster-pdp; Panel B).

@@

First-order PDA showed minimal influence of incoming and outgoing connectivity metrics on
cluster assignment probabilities. Two-way analyses revealed little interaction between depth
and incoming and outgoing connectivity metrics for reef clustering probabilities
(@fig-ts-cluster-pdp-twoway-incoming; @sfig-ts-cluster-pdp-twoway-outgoing). Reefs with a
median depth around 15m may benefit slightly from high incoming and outgoing connectivity,
with higher probability of High cover cluster assignment compared to 15m - depth reefs with
lower connectivity.

@@

Spatially these results indicate that regions of positive carbonate budget maintenance,
under ACCESS-ESM1-5, and high depth levels were concentrated in the Far Northern, offshore
Townsville/Whitsunday, and Mackay/Capricorn management areas (@sfig-depth-map;
@sfig-carb-budget-map). High cover reefs in the Far Northern management area were exposed to
lower mean DHW levels than other reefs under ACCESS-ESM1-5 (@sfig-mean-dhw-map), however
similar spatial patterns were observed across GCMs for reefs clustered into the high cover
cluster (@sfig-gcm-cluster-map).

@@

The authors acknowledge the Traditional Owners of the Sea Country on which the modelling was
conducted for. The authors additionally acknowledge the Bindal, Jagera, and Ngunnawal
Peoples, the lands on which the authors conducted the model development, analysis, and
writing. The authors would like to acknowledge and thank Shaun Wilson for comments and
discussion during manuscript revision. The authors also acknowledge the contributions of
Barbara Robson and Jessica Benthuysen for comments provided early in the manuscript
preparation. The authors would like to additionally acknowledge and thank Shaun Wilson and
Juan Carlos Ortiz for discussions around carbonate budget thresholds.
